### 2023-02-23 ART_Sekrst/Sekrst.tex
Editorial corrections added

\section*{Philosophical essentialism}
An \textit{essential} property of an object is a~property that an object must have, while an \textit{accidental} one is the one the object happens could have, but that it could lack. That is, in modal terms,\footnote{Standard modal characterizations have been disseminated with the works of Ruth Barcan Marcus and Saul Kripke. Kripke's work on semantics is taking the truth of a~formula relative to a~possible world, since its truth value depends on what is true in accessible world. See Barcan Marcus
%\label{ref:RNDktVIAPuvSw}(1993)
\parencite*[][]{marcus_modalities_1993} %
 for a~modality synthesis and Kripke 
%\label{ref:RND50iCIBiyUn}(1972)
\parencite*[][]{kripke_naming_1972} %
 for Kripke semantics.} we are talking about \textit{necessity} and \textit{possibility},\footnote{There are, of course, differences between logical and (meta)physical possibilities. Something might not be a~logical contradiction, but still be (meta)physically impossible, i.e., not conforming to the laws of nature, for example, a~man travelling faster than the speed of light. The exact details of such classifications, especially between physical and metaphysical possibilities, are a~matter of debate.} respectively 
%\label{ref:RNDcC0OGIZEmV}(Robertson Ishii and Atkins, 2020):
\parencite[][]{robertson_ishii_essential_2020}: %
 an object is going to possess the essential property in all possible worlds, but for an accidental one, there is a~possible world in which an object lacks such a~property. \textit{Essentialism} is a~standpoint in which (at least) some objects have (at least some) essential properties 
%\label{ref:RNDS25JSRGYR9}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For example, an essential property of Socrates is to have originated from his parents but is not essential for him to have brown hair. An essential property of a~dog is certainly not brown hair since there are dogs of other colors. In philosophy, some essential properties are not a~matter of much debate. For example, a~dog had to have some biological origin. \textit{Canis canis} is also a~dog. But in order to pursue the matter further, there might be various objections to candidates for essential purposes.\footnote{A~concise description of the debate is provided by Cartwright 
%\label{ref:RNDsX991rAklO}(1968, p.615):
\parencite*[][p.615]{cartwright_remarks_1968}: %
 ``What are the essential attributes of, say, Dancer's Image? No doubt it will be counted essential that he is a~horse and accidental that he was disqualified in this year's Kentucky Derby. But what of the attribute of being male, or of being a~thoroughbred, or of not being a~Clydesdale stallion? Here, I~suppose, essentialists may disagree. Indeed, a~reasonable essentialist might well take the position that these are hard cases that admit of no clear decision.}'' Often, a~dog is considered a~``wolf-like descendant'', where various breeds might not conform to this ideal, along with the notion of ``having an upturning tail''.

@@

The true dichotomy still present is a~wholly different one. In fact, it is the same one that Rene Descartes described half a~millennium ago: do the categories present in my mind have objective validity?\footnote{See Descartes
%\label{ref:RNDahCD1FhxF0}(1641; English translation: 1991)
\parencites*[][]{descartes_renati_1641}[English translation:][]{descartes_meditations_1991} %
 for more details.} The easiest way to a~positive answer is essentialism, which claims that the categories in our minds are formed via the essential properties present in the world. And machine learning, a~new paradigm where machines are finally intelligent enough, is believed by many to show exactly this. If machines can learn the same things we do, then obviously the categories used are not intrinsically human. If machines can learn this by crunching data obtained from the world, then the categories are in fact present in that very data as essential properties. Machine learning is, on this account, simply a~family of algorithms capable of extracting not just information from data, but essential properties as well. As we have shown, this view is wrong, since: (i) this could in theory hold true only for supervised learning, and more importantly (ii) supervised learning is defined via its use of targets or labels which are \textit{man-made}. Since they are man-made, this means that human annotators bring in their categories ``cat/dog'', ``animal/non-animal'', ``happy/sad'', etc., and connect this to actual data, e.g., pixel values, or numeric data. The machine-learning algorithm then extracts this connection and applies it to previously unseen data. But the essential properties are not the ones discovered by the algorithm, they are brought in by human annotators, and do not have to reflect the ``real'' ontology at all. Even in the case of unsupervised learning, the features are being clustered and interpreted by humans, bringing again their own categories into play.

>>>>

\section*{Philosophical essentialism}
An \textit{essential} property of an object is a~property that an object must have, while an \textit{accidental} one is the one the object happens could have, but that it could lack. That is, in modal terms,\footnote{Standard modal characterizations have been disseminated with the works of Ruth Barcan Marcus and Saul Kripke. Kripke's work on semantics is taking the truth of a~formula relative to a~possible world, since its truth value depends on what is true in accessible world. See Barcan Marcus
%\label{ref:RNDktVIAPuvSw}(1993)
\parencite*[][]{marcus_modalities_1993} %
 for a~modality synthesis and Kripke 
%\label{ref:RND50iCIBiyUn}(1972)
\parencite*[][]{kripke_naming_1972} %
 for Kripke semantics.} we are talking about \textit{necessity} and \textit{possibility},\footnote{There are, of course, differences between logical and (meta)physical possibilities. Something might not be a~logical contradiction, but still be (meta)physically impossible, i.e., not conforming to the laws of nature, for example, a~man travelling faster than the speed of light. The exact details of such classifications, especially between physical and metaphysical possibilities, are a~matter of debate.} respectively 
%\label{ref:RNDcC0OGIZEmV}(Robertson Ishii and Atkins, 2020):
\parencite[][]{robertson_ishii_essential_2020}: %
 an object is going to possess the essential property in all possible worlds, but for an accidental one, there is a~possible world in which an object lacks such a~property. \textit{Essentialism} is a~standpoint in which (at least) some objects have (at least some) essential properties 
%\label{ref:RNDS25JSRGYR9}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For example, an essential property of Socrates is to have originated from his parents but is not essential for him to have brown hair. An essential property of a~dog is certainly not brown hair since there are dogs of other colors. In philosophy, some essential properties are not a~matter of much debate. For example, a~dog had to have some biological origin. \textit{Canis canis} is also a~dog. But in order to pursue the matter further, there might be various objections to candidates for essential purposes.\footnote{A~concise description of the debate is provided by Cartwright 
%\label{ref:RNDsX991rAklO}(1968, p.615):
\parencite*[][p.615]{cartwright_remarks_1968}: %
 ``What are the essential attributes of, say, Dancer's Image? No doubt it will be counted essential that he is a~horse and accidental that he was disqualified in this year's Kentucky Derby. But what of the attribute of being male, or of being a~thoroughbred, or of not being a~Clydesdale stallion? Here, I~suppose, essentialists may disagree. Indeed, a~reasonable essentialist might well take the position that these are hard cases that admit of no clear decision.''} Often, a~dog is considered a~``wolf-like descendant'', where various breeds might not conform to this ideal, along with the notion of ``having an upturning tail''.

@@

The true dichotomy still present is a~wholly different one. In fact, it is the same one that René Descartes described half a~millennium ago: do the categories present in my mind have objective validity?\footnote{See Descartes
%\label{ref:RNDahCD1FhxF0}(1641; English translation: 1991)
\parencites*[][]{descartes_renati_1641}[English translation:][]{descartes_meditations_1991} %
 for more details.} The easiest way to a~positive answer is essentialism, which claims that the categories in our minds are formed via the essential properties present in the world. And machine learning, a~new paradigm where machines are finally intelligent enough, is believed by many to show exactly this. If machines can learn the same things we do, then obviously the categories used are not intrinsically human. If machines can learn this by crunching data obtained from the world, then the categories are in fact present in that very data as essential properties. Machine learning is, on this account, simply a~family of algorithms capable of extracting not just information from data, but essential properties as well. As we have shown, this view is wrong, since: (i) this could in theory hold true only for supervised learning, and more importantly (ii) supervised learning is defined via its use of targets or labels which are \textit{man-made}. Since they are man-made, this means that human annotators bring in their categories ``cat/dog'', ``animal/non-animal'', ``happy/sad'', etc., and connect this to actual data, e.g., pixel values, or numeric data. The machine-learning algorithm then extracts this connection and applies it to previously unseen data. But the essential properties are not the ones discovered by the algorithm, they are brought in by human annotators, and do not have to reflect the ``real'' ontology at all. Even in the case of unsupervised learning, the features are being clustered and interpreted by humans, bringing again their own categories into play.

### 2023-02-23 ART_Soloducha/Soloducha.tex
Editorial corrections added

\section*{The Moral Machine project as an implementation of the idea of a~coherent, extrapolated volition}
The Moral Machine project was launched in 2014, being the result of collaboration between several academic centres (Exeter Business School, Massachusetts Institute of Technology, University of British Columbia, Max Planck Institute for Human Development, Toulouse School of Economics). Its aim was to gather via Internet as many opinions on moral dilemmas as possible, using as an example various modifications of the classic trolley model once proposed by Philippa Foot (The issue of the value of the so-called trolley's dilemma for dealing with ethical problems is left here to be discussed in other contexts. Nevertheless, some arguments concerning this problem will be mentioned when discussing the critique of the Moral Machine project).

Using a~special website (http://moralmachine.mit.edu), dilemma scenarios were presented to the public worldwide. The goal of the study was to identify solution patterns that could be used as a~database for implementation in autonomous systems—the reference device in this case was an autonomous car.

@@

There were also attempts to correlate the overall results with a~precise, representative selection of 6 demographic indicators important for the entire survey population - age, education, gender, wealth, religion and political views. The analysis showed no significant differences in the results (the sample is then limited to 492,291 people).

>>>>

\section*{The Moral Machine project as an implementation of the idea of a~coherent, extrapolated volition}
The Moral Machine project was launched in 2014, being the result of collaboration between several academic centres (Exeter Business School, Massachusetts Institute of Technology, University of British Columbia, Max Planck Institute for Human Development, Toulouse School of Economics). Its aim was to gather via Internet as many opinions on moral dilemmas as possible, using as an example various modifications of the classic trolley model once proposed by Philippa Foot.\footnote{The issue of the value of the so-called trolley's dilemma for dealing with ethical problems is left here to be discussed in other contexts. Nevertheless, some arguments concerning this problem will be mentioned when discussing the critique of the Moral Machine project.} 
Using a~special website (http://moralmachine.mit.edu), dilemma scenarios were presented to the public worldwide. The goal of the study was to identify solution patterns that could be used as a~database for implementation in autonomous systems—the reference device in this case was an autonomous car.

@@

There were also attempts to correlate the overall results with a~precise, representative selection of 6 demographic indicators important for the entire survey population--age, education, gender, wealth, religion and political views. The analysis showed no significant differences in the results (the sample is then limited to 492,291 people).

### 2023-02-23 ART_Toscano/Toscano.tex
Editorial corrections added

In the end, cultural mediation---or the ability to think and operate through cultural artifacts---produces historical modes of thinking---i.e. ideologies---and styles of cognition that affect how we learn, think and represent our environment and ourselves. This is what can be described as the notion of distributed cognition
%\label{ref:RNDSg24DwdaBd}(Cole and Engeström, 1993; Gallagher, 2005; 2013).
\parencites[][]{salomon_cultural-historical_1993}[][]{gallagher_how_2005}[][]{gallagher_socially_2013}. %
 Hollan et al. 
%\label{ref:RNDCkpURhADXA}(2000, p.177)
\parencite*[][p.177]{laland_niche_2000} %
 provide a~definition of this process: ``Distributed cognition means more than that cognitive processes are socially distributed across the members of a~group. It is a~broader conception that includes phenomena that emerge in social interactions as well as interactions between people and structure in their environments.'' The notion of distributed cognition has been a~common hypothesis in linguistics and psychology ever since the writings of Vygotsky were published and made available in different translations since the 1970s, but it is by no means the standard model. There are many critics that maintain that, even if aided through different tools, thinking happens basically inside the brain 
%\label{ref:RNDz69682DG3T}(Adams and Aizawa, 2008; 2010; Loh and Kanai, 2016),
\parencites[][]{adams_bounds_2008}[][]{menary_defending_2010}[][]{loh_how_2016}, %
 or they present situations in which thinking is affected from ``outside'' factors (which Clowes terms as ``the impact thesis'' 
%\label{ref:RND8Ryfr6A8K0}(Clowes, 2019)
\parencite[][]{clowes_screen_2019}%
). We will not deal here with those arguments, since they appear to have a~strong need for an essentialist form of conceptualizing cognition.

@@

\myquote{
[S]ymbols are thus not mere expressions, instrumentalities, or correlates of our biological, psychological, and social existence; they are prerequisites of it. Without men, no culture, certainly; but equally, and more significantly, without culture, no men
%\label{ref:RNDKLhGzqGJ1x}(1973, p.49).
\parencite*[][p.49]{}.%
}

@@

A~mental state can best be delineated by the notion of intentionality. Intentionality is a~complex philosophical concept that emerged with Medieval Scholasticism through Medieval Islamic philosophy, but was later retaken and developed in phenomenological circles, starting from the 19\textsuperscript{th} Century. Franz Brentano's work is usually set as a~point of departure for contemporary analyses. In his writings, intentionality is set as an attribute of an individual's mind, which adheres to mental contents, as opposed to attributes of the real world, such as extension and duration, which can be predicated of existing objects. Brentano takes on the discussion from St. Thomas Aquinas, who established that the object which is thought is intentionally in the thinking subject, the object which is loved in the person who loves, the object which is desired in the person desiring, etc. In that sense, intentionality is clearly something that can be predicated of inexistent phenomena, but which has an effect on our own conceptions, desires and beliefs. Brentano
%\label{ref:RNDqcysZzdwoS}(1995, p.68)
\parencite*[][p.68]{brentano_psychology_1995} %
 writes:

@@

}
At this point, intentionality was described as a~clear attribute of mental activity, independent of a~real world, but clearly related to it, and decisive to ascribe it meaning. This trait was important because it offered a~form of cognizing reality without relying on the Kantian formulation that attempted to align (individual) sensations and (social) concepts. In other words, it created a~model where things could be cognized beyond a~thick web of structured epistemological pre-conceptions. This is precisely what encouraged Husserl's enthusiasm, as inscribed in his motto ``Back to the things themselves!'' (\textit{Zurück zu den Sachen selbst}!). For as Merleau-Ponty
%\label{ref:RNDs2qs7wId98}(2005, p.xix)
\parencite*[][p.xix]{merleau-ponty_phenomenology_2005} %
 writes:
 \myquote{
 What distinguishes intentionality from the Kantian relation to a~possible object is that the unity of the world, before being posited by knowledge in a~specific act of identification, is ‘lived' as ready-made or already there.
 }

@@

In the last years, cognitive scientists have developed different models to locate intentionalities via distinctive approaches. The neo-behaviourist Daniel Dennett, for instance, ascribes intentionality to observed rational behavior, and he describes the agent as someone 
\myquote{
who harbors beliefs and desires and other mental states that exhibit intentionality or ‘aboutness', and whose actions can be explained (or predicted) on the basis of the content of these states''
%\label{ref:RNDLWEzQD7Ixq}(Dennett, 1991, p.76).
\parencite[][p.76]{dennett_consciousness_1991}. %
}
 The approach is clear, and amounts to correlating traces to directions and motivations in a~straightforward way. Of course, it is constrained to reading rational behavior and to valuing every action as instrumental to achieve a~specific goal. In contrast, a~neo-pragmatist view 
%\label{ref:RNDSIPGtsCWOd}(Brandom, 1994; 2000; Cash, 2008; 2009)
\parencites[][]{brandom_making_1994}[][]{brandom_articulating_2000}[][]{cash_thoughts_2008}[][]{cash_normativity_2009} %
 proceeds by ascribing intentionality as an explanation and a~specific coupling of action to social norms. As Cash 
%\label{ref:RNDyPGdQx45MP}(2008, p.101)
\parencite*[][p.101]{cash_thoughts_2008} %
 argues:
 \myquote{
 based on the similarity of their movement to the kind of actions, […] would entitle us to ascribe such intentional states as reasons.
 }
  This might be a~key aspect in certain contexts, but it is constrained to knowing what the norms to be applied are, and to evaluating if the ensuing pairing of actions to those norms succeed or not. In that sense, they imply the recognition of patterns, and a~judgment on their application or continuity, but they also underestimate the value of deviance and disregard a~space for individual creativity. Even a~third approach, which we can call a~neo-interactionist perspective, aims at understanding other's intentionalities not by acknowledging or judging their actions, but by understanding actual or potential interactions with others in socially appropriate ways. As Gallagher and Miyahara 
%\label{ref:RND9j2x1Jg8pS}(2012, p.135)
\parencite*[][p.135]{gallagher_neo-pragmatism_2012} %
 write in this account: 
 \myquote{
 we normally perceive another's intentionality in terms of its appropriateness, it's pragmatic and/or emotional value for our particular way of being, constituted by the particular goals or projects we have at the time, or implicit grasp on cultural norms, our social status, and so on, rather than as reflecting inner mental states, or as constituting explanatory reasons for her further thoughts and actions.
 }
 The neo-interactionist perspective certainly rounds up some of the forms in which intentionalities emerge, but they do not completely revoke the previous explanations, and instead helps compile a~catalogue of intentional enactions.

>>>>

In the end, cultural mediation---or the ability to think and operate through cultural artifacts---produces historical modes of thinking---i.e. ideologies---and styles of cognition that affect how we learn, think and represent our environment and ourselves. This is what can be described as the notion of distributed cognition
%\label{ref:RNDSg24DwdaBd}(Cole and Engeström, 1993; Gallagher, 2005; 2013).
\parencites[][]{salomon_cultural-historical_1993}[][]{gallagher_how_2005}[][]{gallagher_socially_2013}. %
 Laland et al. 
%\label{ref:RNDCkpURhADXA}(2000, p.177)
\parencite*[][p.177]{laland_niche_2000} %
 provide a~definition of this process: ``Distributed cognition means more than that cognitive processes are socially distributed across the members of a~group. It is a~broader conception that includes phenomena that emerge in social interactions as well as interactions between people and structure in their environments.'' The notion of distributed cognition has been a~common hypothesis in linguistics and psychology ever since the writings of Vygotsky were published and made available in different translations since the 1970s, but it is by no means the standard model. There are many critics that maintain that, even if aided through different tools, thinking happens basically inside the brain 
%\label{ref:RNDz69682DG3T}(Adams and Aizawa, 2008; 2010; Loh and Kanai, 2016),
\parencites[][]{adams_bounds_2008}[][]{menary_defending_2010}[][]{loh_how_2016}, %
 or they present situations in which thinking is affected from ``outside'' factors (which Clowes
 \parencite*[][]{clowes_screen_2019}%
 terms as ``the impact thesis''%
%\label{ref:RND8Ryfr6A8K0}(Clowes, 2019)
%\parencite[][]{clowes_screen_2019}%
). We will not deal here with those arguments, since they appear to have a~strong need for an essentialist form of conceptualizing cognition.

@@

\myquote{
[S]ymbols are thus not mere expressions, instrumentalities, or correlates of our biological, psychological, and social existence; they are prerequisites of it. Without men, no culture, certainly; but equally, and more significantly, without culture, no men
%\label{ref:RNDKLhGzqGJ1x}(1973, p.49).
\parencite*[][p.49]{geertz_interpretation_1973}.%
}

@@

A~mental state can best be delineated by the notion of intentionality. Intentionality is a~complex philosophical concept that emerged with Medieval Scholasticism through Medieval Islamic philosophy, but was later retaken and developed in phenomenological circles, starting from the 19\textsuperscript{th} century. Franz Brentano's work is usually set as a~point of departure for contemporary analyses. In his writings, intentionality is set as an attribute of an individual's mind, which adheres to mental contents, as opposed to attributes of the real world, such as extension and duration, which can be predicated of existing objects. Brentano takes on the discussion from St. Thomas Aquinas, who established that the object which is thought is intentionally in the thinking subject, the object which is loved in the person who loves, the object which is desired in the person desiring, etc. In that sense, intentionality is clearly something that can be predicated of inexistent phenomena, but which has an effect on our own conceptions, desires and beliefs. Brentano
%\label{ref:RNDqcysZzdwoS}(1995, p.68)
\parencite*[][p.68]{brentano_psychology_1995} %
 writes:

@@

}
At this point, intentionality was described as a~clear attribute of mental activity, independent of a~real world, but clearly related to it, and decisive to ascribe it meaning. This trait was important because it offered a~form of cognizing reality without relying on the Kantian formulation that attempted to align (individual) sensations and (social) concepts. In other words, it created a~model where things could be cognized beyond a~thick web of structured epistemological pre-conceptions. This is precisely what encouraged Husserl's enthusiasm, as inscribed in his motto ``Back to the things themselves!'' (\textit{Zurück zu den Sachen selbst!}). For as Merleau-Ponty
%\label{ref:RNDs2qs7wId98}(2005, p.xix)
\parencite*[][p.xix]{merleau-ponty_phenomenology_2005} %
 writes:
 \myquote{
 What distinguishes intentionality from the Kantian relation to a~possible object is that the unity of the world, before being posited by knowledge in a~specific act of identification, is ‘lived' as ready-made or already there.
 }

@@

In the last years, cognitive scientists have developed different models to locate intentionalities via distinctive approaches. The neo-behaviourist Daniel Dennett, for instance, ascribes intentionality to observed rational behavior, and he describes the agent as someone 
\myquote{
who harbors beliefs and desires and other mental states that exhibit intentionality or ‘aboutness', and whose actions can be explained (or predicted) on the basis of the content of these states
%\label{ref:RNDLWEzQD7Ixq}(Dennett, 1991, p.76).
\parencite[][p.76]{dennett_consciousness_1991}. %
}
 The approach is clear, and amounts to correlating traces to directions and motivations in a~straightforward way. Of course, it is constrained to reading rational behavior and to valuing every action as instrumental to achieve a~specific goal. In contrast, a~neo-pragmatist view 
%\label{ref:RNDSIPGtsCWOd}(Brandom, 1994; 2000; Cash, 2008; 2009)
\parencites[][]{brandom_making_1994}[][]{brandom_articulating_2000}[][]{cash_thoughts_2008}[][]{cash_normativity_2009} %
 proceeds by ascribing intentionality as an explanation and a~specific coupling of action to social norms. As Cash 
%\label{ref:RNDyPGdQx45MP}(2008, p.101)
\parencite*[][p.101]{cash_thoughts_2008} %
 argues:
 \myquote{
 based on the similarity of their movement to the kind of actions, […] would entitle us to ascribe such intentional states as reasons.
 }
  This might be a~key aspect in certain contexts, but it is constrained to knowing what the norms to be applied are, and to evaluating if the ensuing pairing of actions to those norms succeed or not. In that sense, they imply the recognition of patterns, and a~judgment on their application or continuity, but they also underestimate the value of deviance and disregard a~space for individual creativity. Even a~third approach, which we can call a~neo-interactionist perspective, aims at understanding other's intentionalities not by acknowledging or judging their actions, but by understanding actual or potential interactions with others in socially appropriate ways. As Gallagher and Miyahara 
%\label{ref:RND9j2x1Jg8pS}(2012, p.135)
\parencite*[][p.135]{gallagher_neo-pragmatism_2012} %
 write in this account: 
 \myquote{
 we normally perceive another's intentionality in terms of its appropriateness, it's pragmatic and/or emotional value for our particular way of being, constituted by the particular goals or projects we have at the time, or implicit grasp on cultural norms, our social status, and so on, rather than as reflecting inner mental states, or as constituting explanatory reasons for her further thoughts and actions.
 }
 The neo-interactionist perspective certainly rounds up some of the forms in which intentionalities emerge, but they do not completely revoke the previous explanations, and instead helps compile a~catalogue of intentional enactions.

### 2023-02-23 EDI_Krzanowski/Krzanowski.tex
Editorial corrections added

Some of the papers collected in this edition of ZFN were presented at the Philosophy in Informatics VI: Frontiers of Philosophy of Computing and Information conference held on December 16–17 and organized by the Polish Academy of Arts and Sciences (PAU) 2021. We would like to thank Gordana Dodig-Crnkovic for supporting this conference and contributing to this special edition.

>>>>

Some of the papers collected in this edition of ZFN were presented at the Philosophy in Informatics VI: Frontiers of Philosophy of Computing and Information conference held on December 16–17, 2021, and organized by the Polish Academy of Arts and Sciences (PAU). We would like to thank Gordana Dodig-Crnkovic for supporting this conference and contributing to this special edition.
